Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-08-18 response.tex

\comment{(3) In the description of the simulation results, the eBayes and ShrinkBayes methods display very similar performance. The distinction between the results on the Figures is quite minimal. The authors should attempt to quantify the difference in performance in a single metric. For instance, in Figure 1, take the average percent difference in TPR across the FPRs that were used to generate the plot.}

Reviewer \#4

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\comment{2. The model appears reasonable and is adapted from on used with microarrays.  To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.}

\comment{a. How robust do you feel your method is?  There is no discussion along these lines.}

\comment{b.      Why is there more of an improvement over ShrinkBayes with more replicates?  Is it because of the method used to calculate the posterior probabilities?  I'd like to see some discussion on this.}

\comment{c.      The ROC curves are not particularly impressive.  Even with 16 reps, the true positive rate is below 50\% (controlling for the false positive rate).  The real data example included only 4 reps, which is far more likely a scenario.  Very little discussion is devoted to this.}

\comment{3.      Minor edits or comments}

\comment{a.      Start of Section 2 on page 2  :  Replace "We considering" with just "Consider"}

\response{Fixed.}

\comment{b.      Page 3: "The parental averages and overdispersion parameters are assumed to follow normal distributions, i.e.,"}

\response{Fixed.}

\comment{c.      Right before Section 2.2…it is assumed apriori independence, correct?}

\response{Yes, added \emph{a priori}.}

\comment{d.      Page 4: You have a two-step approach where you estimate the parameters of interest to get estimates of the hyperparameters and then restimate.  How much change is there is these estimates?  Why aren't the methods of McCarthy used as a competitor?}

\comment{e.      Page 11 "These figures show marked departures from ……"}

\response{Fixed.}

\comment{f.      Page 11 : Does it really matter that the posterior is different from the prior?  One often assume aprior independence but expects there to be posterior correlation.  Perhaps I'm just missing something in this discussion.}

### 2015-08-18 response.tex

\documentclass{article}

\usepackage{color}

\usepackage[parfill]{parskip}

\newcommand{\comment}[1]{\textit{#1}}
\newcommand{\response}[1]{#1}
\newcommand{\todo}[1]{{\color{red} #1}}

@@

\begin{document}

Editor comments:

\comment{The proposed model appears to be an extension of Ji et al. (2014) albeit with a differently specified likelihood function (negative binomial or Poisson-gamma) as perceived to be more appropriate for count data relative to the Gaussian likelihood function pursued in Ji et al. (2014). Both reviewers indicate that they are little underwhelmed by the performance of your proposed procedure relative to some of the existing methods that you compare against (i.e., ShrinkBayes, edgeR and baySeq).  Furthermore, they seem somewhat concerned about an ‘apples vs. oranges’ types of comparisons since the proposed hypothesis testing strategy in the proposed method is really not an option in the other methods.}

\comment{Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}

@@

\comment{I realize that Ji et al.’s (2014) work was intended for microarray data, or more specifically Gaussian likelihoods, but there has been some work that suggests that Gaussian likelihood specifications in MAANOVA may, oddly enough, have better properties for the analysis of RNA-seq data than more popular methods specifically intended for count data (edgeR and DESeq).  See Reeb and Steibel (2013) and also comments from one reviewer along these lines.  Since Ji et al.’s work also derives from ISU, have you assessed its performance relative to your proposed method on RNA-seq data?}

% Email sent to Tieming Ji to see if she has the script.

\comment{One seemingly glaring omission from this work is any comparative assessment of the proposed versus other methods (e.g. ShrinkBayes) on the actual maize dataset in Section 4.  Why is that missing?}

\comment{The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}

\comment{Page 1.  I was instructed that heterosis is merely characterized by the hybrid performance not being equal to the average of the two parental lines.  So in your very first sentence, I might write “Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the average phenotypes of their parents.” i.e., I might be only interested in the $\delta_g$’s. However, your later characterization of heterosis is really something called overdominance to me…where the hybrid performance exceeds the performance of either parent.  I wonder if you could then be more specific in your definitions in your first sentence…and also in your third sentence in this particular paragraph. lines 48-59.  Perhaps this was already captured well in the Ji et al. (2014) paper but why not simply test heterosis as the difference between the hybrid and the average parental performance (i.e. $\mu_3 - 0.5 \mu_1 - 0.5 \mu_2$).  Again, I think your issue pertains really to an overdominance test of some sort.}

\response{Yes, overdominance is the scenario were are interested in. We have added the term in the introduction as well as in Section 2.3 to make this clear.}

@@

\comment{Page 3. Bottom.  Shouldn’t you at least provide a justification when one would use a normal prior and when one would or should use a Laplace prior?}

\todo{currently have normal versions running}

\comment{Page 4, bottom half.  The strategy described here, also somewhat attested to by a reviewer, is a little disconcerting with respect to the “double estimation” of gene-specific parameters…first to estimate they hyperparameters and then again conditioned on the estimates of these hyperparameters.  Perhaps you should characterize that these second-stage gene-specific estimates are shrunk relative to those estimates used to estimate the hyperparameters and that they then have lead to predictions with better properties?  Ideally, I would have liked to see these hyperparameters be also formally estimated together with the gene specific parameters within one MCMC analyses…authors should address why they didn’t do this.}

\response{We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.}

@@

\comment{Page 5, lines 37-47.  Since you’re doing MCMC, why not simply tabulate the number of times samples of $\mu_3$ exceeds either $\mu_1$ or $\mu_2$….and/or the number of times that $\mu_3$ is less than $\mu_1$ or $\mu_2$.  I suspect that what your test is really doing….in which case it might be useful to explicitly state this equivalence as such.}

\response{Yes, that is what the test is doing. We have now stated the case for LPH, i.e. ``$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$).''

@@

\todo{currently have normal versions running}

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\todo{No, because we did not simulate from our model since we wanted to maintain the inherent structure of the gene-specific parameters. Perhaps include a sentence based on the other simulation study. Perhaps say that we do not believe they are estimated very well.}

@@

\comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}

\response{It is not necessary, but it is sufficient. If the parents are the same, then the hybrid must have overdominance.}

@@

\comment{Page 9, lines 23-25.  In tandem with another reviewer’s comments, the real question is why?  Is it because of the suboptimality of the test that you contrived for ShrinkBayesA?}

\todo{Why does ShrinkBayes perform worse? When the normal results are done, we can speak more directly to this.}

@@

\comment{Page 10, Line 29.  Don’t you mean “nadir” rather than peak?}

\response{No, although we do mean ridge rather than peak, i.e. the count is high for effect size of zero and probabilities below 0.5.}

@@

\comment{Page 11 line 39.  Is it really “independence” or nearly non-identifiability that is the issue?}

\todo{??}

\comment{Page 11 last paragraph.  So this then begs an assessment of how good/poor these hyperparameter estimates were.}

\todo{Add discussion of quality of hyperparameter estimates.}

@@

Reviewer \#1

\comment{The authors develop a hierarchical negative binomial model for drawing inferences from a composite null hypothesis, that tests gene expression heterosis. Simulations are performed to describe the performance of this new method along with a data analysis demonstrating an application to real data. This manuscript represents a well-conceived and executed methodological development in the statistical analysis of RNA-Seq gene expression data. However, the performance of the new method is underwhelming compared to existing methods, tempering my enthusiasm for the likely impact of this paper.}

\comment{(1) In the simulations, the authors should also compare the eBayes method to the method of Ji et al. (2014). Although developed for continuous microarray data, the authors could apply a simple transformation to the count data in order to make them continuous. I would be curious to know whether the eBayes method out-performs the method of Ji et al. (2014) on the transformed data.}

\todo{Will run when we get the code. If we don't get the code, will figure out how to run it. Ideally it would be nice to see the eBayes version as well as the approximation in Ji et al.}

@@

\comment{(2) The authors should provide more details on the computational price of the eBayes method? How long (in comparison to edgeR, baySeq, ShrinkBayes) did the eBayes method take to run? Are these methods comparable in terms of computational cost?}

\todo{We will do this.}

\comment{(3) In the description of the simulation results, the eBayes and ShrinkBayes methods display very similar performance. The distinction between the results on the Figures is quite minimal. The authors should attempt to quantify the difference in performance in a single metric. For instance, in Figure 1, take the average percent difference in TPR across the FPRs that were used to generate the plot.}

\todo{I think the main point is that they aren't very different. Somehow I am not able to communicate that the model is novel and that here are a couple of different estimation approaches.}

@@

Reviewer \#4

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\todo{Dan? Even though some genes may be good candidates. An exploratory analysis to suggest other genes seems relevant.}

@@

\comment{2. The model appears reasonable and is adapted from on used with microarrays.  To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.}

\comment{a. How robust do you feel your method is?  There is no discussion along these lines.}

\todo{Good question. Although we suggest the gene-specific parameter distributions do not fit very well, how much better do we really think we can get if we model it well?}

@@

\comment{b.      Why is there more of an improvement over ShrinkBayes with more replicates?  Is it because of the method used to calculate the posterior probabilities?  I'd like to see some discussion on this.}

\todo{Once normal runs are completed, we can comment on this more.}

@@

\comment{c.      The ROC curves are not particularly impressive.  Even with 16 reps, the true positive rate is below 50\% (controlling for the false positive rate).  The real data example included only 4 reps, which is far more likely a scenario.  Very little discussion is devoted to this.}

\todo{Yes, but we haven't created artificially large signals.}

@@

\comment{c.      Right before Section 2.2…it is assumed apriori independence, correct?}

\response{Yes, added \emph{a priori}.}

\comment{d.      Page 4: You have a two-step approach where you estimate the parameters of interest to get estimates of the hyperparameters and then restimate.  How much change is there is these estimates?  Why aren't the methods of McCarthy used as a competitor?}

\todo{Comment on the change in gene-specific parameter estimates, e.g. shrinkage.}

@@

\comment{e.      Page 11 "These figures show marked departures from ……"}

\response{Fixed.}

\comment{f.      Page 11 : Does it really matter that the posterior is different from the prior?  One often assume aprior independence but expects there to be posterior correlation.  Perhaps I'm just missing something in this discussion.}

\todo{Yes, because we are only estimating the parameters of assumed marginal distributions that are assumed independent. If we were performing a fully Bayesian analysis, then we could learn about the joint distribution of the gene-specific parameters.}

### 2015-08-18 ebayes_rnaseq_heterosis.tex
responses to all comments, I htink

Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the phenotypes of their parents.  The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.  One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.  Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}. One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.

@@

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene expression heterosis. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis. The data and code necessary to perform this analysis are available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

%We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

@@

To perform this Bayesian analysis, we used the statistical software \Stan{} which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}. We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference. We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}. If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burnin and inference. We continued this restarting and doubling until we obtained minimum ESS greater than 1000 for all parameters.

@@

For a specific gene $g$, low-parent heterosis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

% \subsection{Implementation in \Stan{}}

%To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous. With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in \RStan{}. Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.

@@

{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for extreme heterosis.

@@

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}
The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero. Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.

>>>>

Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the average phenotypes of their parents.  The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production. Overdominance is a particularly extreme case of heterosis where the hybrid phenotype lies outside the range of the parental phenotypes. One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.  Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}. One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.

@@

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of overdominance. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis. The data and code necessary to perform this analysis are available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

@@

Initial attempts to estimate the models above via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains. For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors that suggested we would need to run at least ten times as long to obtain convergence. Although there are certainly improvements that could be made to decrease computational costs and improve mixing, we opted for an empirical Bayes approach. This approach may be a reasonable approximation to a fully Bayesian approach when estimating models with large numbers of genes as the posterior distributions for the hyperparameters may be tightly peaked.

We categorize the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

@@

To perform this Bayesian analysis, we used the statistical software \Stan{} \citep{stan-software:2014} executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}. \Stan{} implements a variant of MCMC called Hamiltonion Monte Carlo \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}. We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference. We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}. If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burnin and inference. We continued this restarting and doubling until we obtained minimum ESS greater than 1000 for all parameters.

@@

In the maize context that motivates this work, we are particularly interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$), and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.

@@

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}
The figure shows a ridge at an effect size of zero for probabilities below 0.5. Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.
